# Introduction

```{r}
#| echo: false

source("_common.R")
```

## **Prerequisites**

We’ve made a few assumptions about what you already know to get the most out of this book. You should have some basic knowledge about medical statistics, and it’s helpful if you have some basic R programming experience already.

You need some things to run the code in this book: R, RStudio and some preinstalled R packages. Packages are the fundamental units of reproducible R code. They include reusable functions, documentation that describes how to use them, and sample data.

### **R**

To download R, go to CRAN, the **c**omprehensive **R** **a**rchive **n**etwork, [https://cloud.r-project.org](#0). A new major version of R comes out once a year, and there are 2-3 minor releases each year. It’s a good idea to update regularly. Upgrading can be a bit of a hassle, especially for major versions that require you to re-install all your packages, but putting it off only makes it worse. We recommend R 4.4.1 or later for this book.

### **RStudio**

RStudio is an integrated development environment, or IDE, for R programming, which you can download from [https://posit.co/download/rstudio-desktop/](#0). RStudio is updated a couple of times a year, and it will automatically let you know when a new version is out, so there’s no need to check back. It’s a good idea to upgrade regularly to take advantage of the latest and greatest features. For this book, make sure you have at least RStudio 2024.04.2.

## **Running R code**

The previous section showed you several examples of running R code. The code in the book looks like this:

```{r}
1 + 2
```

If you run the same code in your local console, it will look like this:

``` yacc
> 1 + 2
[1] 3
```

There are two main differences. In your console, you type after the `>`, called the **prompt**; we don’t show the prompt in the book. In the book, the output is commented out with `#>`; in your console, it appears directly after your code. These two differences mean that if you’re working with an electronic version of the book, you can easily copy code out of the book and paste it into the console.

Throughout the book, we use a consistent set of conventions to refer to code:

-   Functions are displayed in a code font and followed by parentheses, like `sum()` or `mean()`.

-   Other R objects (such as data or function arguments) are in a code font, without parentheses, like `flights` or `x`.

-   Sometimes, to make it clear which package an object comes from, we’ll use the package name followed by two colons, like `dplyr::mutate()` or `nycflights13::flights`. This is also valid R code.

-   To improves readability, variable names and function names are named using snake case.

## **Colophon**

The book is written by Quarto, an online version of it is available at [https://qbgaoo.github.io/r4ms/](#0). It will continue to evolve in between reprints of the physical book. The source of the book is available at <https://qbgaoo.github.io/r4ms/>.
