# Normal distribution and medical reference range

```{r}
#| echo: false

source("_common.R")
```

The normal distribution, also known as the Gaussian distribution, is one of the most important probability distributions in statistics. It is characterized by its "bell-shaped" curve, which is symmetric about the mean. The normal distribution is used to model a wide range of real-world phenomena, including heights and weights of adults, blood pressures, reaction times in psychological experiments, and more.

## Prerequisite

Setting up the required R packages in this chapter.

```{r}
#| message: false

library(tidyverse)
library(nortest)
library(scales)
```

## Normal distribution

The probability density function of a normal distribution is given by the formula:

$$f(x) = \frac{1}{\sqrt{2\pi\sigma^2}} e^{-\frac{(x - \mu)^2}{2\sigma^2}}$$

where $x$ is the variable, $\mu$ is the mean and $\sigma$ is the standard deviation of the distribution.

```{r}
tibble(
  x  = seq(-4, 4, length.out = 150),
  y1 = dnorm(x, mean = 0,   sd = 1),
  y2 = dnorm(x, mean = -1,  sd = 1)
) |> 
  pivot_longer(
    cols      = contains("y"),
    values_to = "density",
    names_to  = "group"
  ) |> 
  ggplot(aes(x = x, y = density, linetype = group)) +
  geom_line() +
  scale_linetype_manual(
    name   = "",
    values = c("solid", "dotdash"),
    labels = c("mean = 0", "mean = -1")
  ) +
  theme(
    axis.text   = element_text(size = 12),
    axis.title  = element_text(size = 12),
    legend.text = element_text(size = 12)
  )
```

```{r}
tibble(
  x  = seq(-4, 4, length.out = 150),
  y1 = dnorm(x, mean = 0, sd = 0.5),
  y2 = dnorm(x, mean = 0, sd = 1),
  y3 = dnorm(x, mean = 0, sd = 1.5)
) |> 
  pivot_longer(
    cols      = contains("y"),
    values_to = "density",
    names_to  = "group"
  ) |> 
  ggplot(aes(x = x, y = density, linetype = group)) +
  geom_line() +
  scale_linetype_manual(
    name   = "",
    values = c("solid", "dotdash", "dashed"),
    labels = c("sd = 0.5", "sd = 1", "sd = 1.5")
  ) +
  theme(
    axis.text   = element_text(size = 12),
    axis.title  = element_text(size = 12),
    legend.text = element_text(size = 12)
  )
```

```{r}
df <- tibble(
  x       = seq(-4, 4, length.out = 600),
  density = dnorm(x, mean = 0,   sd = 1)
) 

df |> 
  ggplot(aes(x = x, y = density)) +
  geom_line() +
  geom_area(
    data = filter(df, x >= 1.96),
    aes(x = x, y = density),
    alpha = 0.5
  ) + 
  geom_area(
    data = filter(df, x <= -1.96),
    aes(x = x, y = density),
    alpha = 0.5
  ) + 
  scale_x_continuous(
    breaks = c(-1.96, 0, 1.96),                    
    labels = expression(mu-1.96*sigma, mu, mu+1.96*sigma)) +
  annotate(
    "text", x = -2.6, y = 0.04, 
    label = percent(pnorm(-1.96), accuracy = 0.1)
  ) +  
  annotate(
    "text", x = 2.6, y = 0.04, 
    label = percent(1 - pnorm(1.96), accuracy = 0.1)
  ) +  
  theme(
    axis.text   = element_text(size = 12),
    axis.title  = element_text(size = 12),
    legend.text = element_text(size = 12)
  )
```

Area of the shaded tails

```{r}

```

## 

::: example
Some researchers used a random sampling method to examine the red blood cell count of 55 normal adults. The measuring results are saved in a `.csv` file. Please analyze its normality.
:::

<a href="datasets/ex03-01.csv" download="ex03-01.csv" class="btn btn-success"> <i class="bi bi-database-fill-down"></i> Download data </a>

You can click on the download button above to download and the save it in your own folder. Here we import the data file into R and assign to a tibble named `rbc`.

```{r}
rbc <- read_csv("datasets/ex03-01.csv")
```

Test of normality is used to determine if a data is well-modeled by a normal distribution. This is an important assumption for many statistical tests. Here are some common tests of normality:

-   Shapiro-Wilk test: Best for small to medium-sized data.
-   Shapiro-Francia test: A variation of the Shapiro-Wilk test. It is generally more appropriate for dealing with larger sample size data compared to the Shapiro-Wilk test, particularly for data that is expected to be normally distributed.
-   Anderson-Darling test: Gives more weight to the tails of the distribution.
-   Q-Q Plot: Always useful as a supplementary method.

Here we perform the Shapiro-Wilk test of normality using `shapiro.test()`, which lies in `stats` package. The difference is Shapiro-Francia test and Anderson-Darling test are performed by `sf.test()` and `ad.test()`, respectively. Both of them come from the `nortest` package, which need to be installed before used.

::: panel-tabset
## Shapiro-Wilk

```{r}
rbc |> 
  pull() |> 
  shapiro.test()
```

## Shapiro-Francia

```{r}
rbc |> 
  pull() |> 
  sf.test()
```

## Anderson-Darling

```{r}
rbc |> 
  pull() |> 
  ad.test()
```
:::

-   If the p-value is greater than 0.05, the data is normally distributed (fail to reject the null hypothesis).
-   If the p-value is less than 0.05, the data is not normally distributed (reject the null hypothesis).

In addition to formal tests, visual inspection of the data can be very informative. You can create a Q-Q Plot (Quantile-Quantile Plot) using `qqnorm()` function. More, you can add a line to a theoretically normal quantile-quantile plot by `qqline()`, which passes through the first and third quartiles by default.

```{r}
rbc |> 
  pull() |> 
  qqnorm(main = "", datax = T)

rbc |> 
  pull() |> 
  qqline(datax = T)
```

Alternatively you can use the `ggplot2` package to create a Q-Q plot, which has more customization and flexibility. Here is an example.

```{r}
rbc |>
  ggplot(aes(sample = rbc)) +
  geom_qq(shape = 1, size = 2.3) +
  geom_qq_line() +
  labs(x = "Theoretical Quantiles", y = "Sample Quantiles") +
  coord_flip() +
  theme(
    axis.text  = element_text(size = 12),
    axis.title = element_text(size = 12)
  )
```

-   If the data points fall approximately along the reference line in the Q-Q plot, the data is likely normally distributed.
-   Significant deviations from the line indicate departures from normality.

## Medical reference range

A medical reference range is a set of values that is used by healthcare providers to interpret laboratory test results. These ranges are typically established by testing a large group of healthy individuals and determining the normal distribution of values within that population. The reference range provides a context for understanding whether a individual’s test results fall within a normal, expected range or if they indicate a potential health issue.

```{r}
med_ref_range <- rbc |> 
  pull() |> 
  quantile(probs = c(0.025, 0.975)) |> 
  round(digits = 2) |> 
  print()
```

The result indicates that the 95% medical reference range for the maximum heart rate in normal individuals is: `r paste0("(", med_ref_range[1], ", ", med_ref_range[2], ")")`.
